Phenotypic evolution is more constrained in simpler food webs

### 2018-11-22 manuscript.Rmd
incorporated Jordi's comments on Intro.

The adaptive landscape provides a powerful framework for understanding how natural selection has shaped the evolution of biodiversity ---from genes, to phenotypes, to species [@Wright1931; @Simpson1944; @Arnold2001]. More than a metaphor, the adaptive landscape links quantitative genetic and phenotypic variation to evolution by natural selection [@Lande1979; @Arnold1984applications; @Arnold1984theory]. Ecological interactions often play a key role in shaping adaptive landscapes, as evidenced by the role of antagonistic [@Schluter2000; @Abrams2000] and mutualistic [@Bronstein2006] interactions in driving evolutionary change.<!--, often play a key role in shaping the adaptive landscape and phenotypic evolution in multiple traits [; ;]. For example, competition for shared prey in threespine stickleback has driven their adaptive divergence into ecotypes with specialized body and gill morphologies [@Schluter2000, other papers]. Similarly, the elongated, red corollas of many flowering species have evolved in response to the efficient pollination of hummingbirds [cite].--> Although there is clear evidence that pairwise interactions can shape the adaptive landscape, we also know that most species interact with multiple species in a community context [@Darwin1859]. Understanding how the adaptive landscape is shaped by community context represents a major frontier for evolutionary ecologists [@McPeek2017; @terHorst2018]. Resolution on this front is urgently needed though, given the rapid impacts of climate change on ecological communities [@Scheffers2016].

Ecological networks, such as a food web describing who eats whom, provide an explicit representation of the community context. These networks describe the interdependency of populations within an ecological community, providing an effective framework for predicting how the strength and organization of ecological interactions (network structure) affect population and community dynamics [@Bascompte2014; @McCann2012]. Nevertheless, these networks often lack quantitative information on how ecological interactions influence selection on phenotypic traits. Thus, it remains unclear how network structure shapes adaptive landscapes of interacting populations. This represents a major barrier in our current ability to predict how changes in community context will affect phenotypic evolution.

Here, we integrate ecological networks and adaptive landscapes to understand how community context constrains evolutionary change [@Arnold1992]. Different aspects of evolutionary constraints can be inferred by quantifying the slope and curvature of the adaptive landscape. For example, the slope is determined by directional selection gradients acting on each phenotypic trait and influences the trajectory of evolutionary change [@Lande1979;@Arnold1992]. Evolutionary trajectories become more constrained with an increase in the number of traits under selection, as this diminishes the number of optimal solutions [@Arnold2003]. The curvature of the adaptive landscape can also constrain evolution through its indirect effect on genetic constraints [@Arnold1992; @Hansen2008]. Genetic constraints are largely governed by a population's **G**-matrix ---the additive genetic variances and covariances between traits [@Hansen2008]. In general, genetic constraints will increase with the number of traits under directional or stabilizing selection, as this will decrease the additive genetic variance in those traits [@Hansen2008]. Genetic constraints may also increase with the number of trait combinations under correlational selection, as this type of selection decreases the evolutionary independence of traits [@Hansen2008]. If we want to predict how community context constrains evolutionary change, we must understand how ecological networks shape the adaptive landscape.

To examine the effect of network structure on adaptive landscapes, we conducted a field experiment that manipulated the complexity of trophic interactions (hereafter food-web complexity) associated with an abundant insect herbivore (*Iteomyia salicisverruca*; Family Cecidomyiidae)(fig. \ref{fig:Conceptual}). The larvae of this herbivore induce tooth-shaped galls when they feed on the developing leaves of willow trees [*Salix* sp., @Russo2006]. These galls protect larva from attack by generalist predators (e.g. ants, spiders), but they suffer high mortaltiy from egg and larval parasitoids [@Barbour2016]. We manipulated food-web complexity by either excluding the guild of larval parasitoids (simple food web) or allowing both egg and larval parasitoids to impose selection on gall midge traits (complex food web; fig. \ref{fig:Conceptual}). Food-web complexity may influence the adaptive landscape in a number of ways. For example, if each parasitoid imposes directional selection on a different trait, then more complex food webs may constrain evolution by increasing the number of traits under selection. Alternatively, if parasitoids impose selection on different values of a trait, then their selective effects would cancel each other out in more complex food webs. To examine these different possibilities (among others), we applied modern statistical methods to quantify the effect of food-web complexity on the slope and curvature of the gall midge's adaptive landscape. Taken together, our study gives insight to how local extinctions, and concomitant decreases in food-web complexity, may constrain the evolution of interacting populations.

>>>>

The adaptive landscape provides a powerful framework for understanding how natural selection has shaped the evolution of biodiversity ---from genes to phenotypes to species [@Wright1931; @Simpson1944; @Arnold2001]. More than a metaphor, the adaptive landscape links quantitative genetic and phenotypic variation to evolution by natural selection [@Lande1979; @Arnold1984applications; @Arnold1984theory]. Ecological interactions often play a key role in shaping adaptive landscapes, as evidenced by the role of antagonistic [@Schluter2000; @Abrams2000] and mutualistic [@Bronstein2006] interactions in driving evolutionary change.<!--, often play a key role in shaping the adaptive landscape and phenotypic evolution in multiple traits [; ;]. For example, competition for shared prey in threespine stickleback has driven their adaptive divergence into ecotypes with specialized body and gill morphologies [@Schluter2000, other papers]. Similarly, the elongated, red corollas of many flowering species have evolved in response to the efficient pollination of hummingbirds [cite].--> Although there is clear evidence that pairwise interactions can shape the adaptive landscape, we do not know how the adaptive landscape is shaped by the community context [@McPeek2017; @terHorst2018]. Resolution on how the community context shapes phenotypic evolution is urgently needed though, given the rapid impacts of climate change on ecological communities [@Scheffers2016].

Ecological networks, such as food webs describing who-eats-whom in ecological communities, provide an explicit representation of the community context [@Bascompte2014; @McCann2012].<!--These networks describe the interdependency of populations within an ecological community, providing an effective framework for predicting how the strength and organization of ecological interactions (network structure) affect population and community dynamics [@Bascompte2014; @McCann2012]. Nevertheless, these networks often lack quantitative information on how ecological interactions influence selection on phenotypic traits. Thus, it remains unclear how network structure shapes adaptive landscapes of interacting populations. This represents a major barrier in our current ability to predict how changes in community context will affect phenotypic evolution.--> Here, we integrate ecological networks and adaptive landscapes to understand how community context constrains evolutionary change [@Arnold1992]. Different aspects of evolutionary constraints can be inferred by quantifying the slope and curvature of the adaptive landscape. For example, the slope is determined by directional selection gradients acting on each phenotypic trait and influences the trajectory of evolutionary change [@Lande1979;@Arnold1992]. Evolutionary trajectories become more constrained with an increase in the number of traits under selection, as this diminishes the number of optimal solutions [@Arnold2003]. The curvature of the adaptive landscape can also constrain evolution through its indirect effect on genetic constraints [@Arnold1992; @Hansen2008]. Genetic constraints are largely governed by a population's **G**-matrix ---the additive genetic variances and covariances between traits [@Hansen2008]. In general, genetic constraints will increase with the number of traits under directional or stabilizing selection, as this will decrease the additive genetic variance in those traits [@Hansen2008]. Genetic constraints may also increase with the number of trait combinations under correlational selection, as this type of selection decreases the evolutionary independence of traits [@Hansen2008]. If we want to predict how community context constrains evolutionary change, we must understand how ecological networks shape the adaptive landscape.

Global change is simplifying the structure of ecological communities, which may influence evolutionary constraints in a number of ways. For example, in a multitrophic community or food web, if different consumers impose directional selection on different traits of a shared resource, then more complex food webs may constrain evolution by increasing the number of traits under selection. Alternatively, if consumers impose selection on different values of a trait, then their selective effects would cancel each other out in more complex food webs. To examine these different possibilities (among others), <!--To examine the effect of network structure on adaptive landscapes, -->we conducted a field experiment that manipulated the complexity of trophic interactions (hereafter food-web complexity) associated with an abundant insect herbivore (*Iteomyia salicisverruca*; Family Cecidomyiidae)(fig. \ref{fig:Conceptual}). The larvae of this herbivore induce tooth-shaped galls when they feed on the developing leaves of willow trees [*Salix* sp., @Russo2006]. These galls protect larva from attack by generalist predators (e.g. ants, spiders), but they suffer high mortaltiy from egg and larval parasitoids [@Barbour2016]. We manipulated food-web complexity by either excluding the guild of larval parasitoids (simple food web) or allowing both egg and larval parasitoids to impose selection on gall midge traits (complex food web; fig. \ref{fig:Conceptual}). <!--Food-web complexity may influence the adaptive landscape in a number of ways. For example, if each parasitoid imposes directional selection on a different trait, then more complex food webs may constrain evolution by increasing the number of traits under selection. Alternatively, if parasitoids impose selection on different values of a trait, then their selective effects would cancel each other out in more complex food webs. To examine these different possibilities (among others), -->We then applied modern statistical methods to quantify the effect of food-web complexity on the slope and curvature of the gall midge's adaptive landscape. Taken together, our study gives insight to how local extinctions, and concomitant decreases in food-web complexity, may constrain the evolution of interacting populations.
